---
title: "Reproducibility"
subtitle: "(& Machine Learning)"
author: "Thomas Wise"
format: 
  revealjs:
    smaller: true
    theme: dark
editor: visual
---

## Session Objectives

-   Understand the importance of Reproducibility within Statistical Analysis
-   Understand the importance of Reproducibility within Machine Learning
-   Be able to apply measures of Reproducibility within Coding

## Today's Plan

-   Section 1: The Importance of Reproducibility
-   Section 2: Introduction to Machine Learning
-   Exercise 1: Applying Reproducibility, through Machine Learning Preparation
-   Break
-   Section 3: Developing Machine Learning Models
-   Exercise 2: Applying Machine Learning Models

# But first - Setup!

Similar to last week, please take a moment to ensure that the tidy models package(s) are loaded in correctly!

# Section 1: The Importance of Reproducibility

## Why is Reproducibility Important

-   When conducting statistical analysis, the only variable is the input.
-   We expect that output to change, when the input changes.
-   But we want our statistical analysis to be repeatable, so that there is a clear audit trail.

## How to achieve Reproducibility

-   At their heart, computers do not like randomness, meaning we can exploit that to create Reproducible behaviour despite randomness.

For example:

-   If you run the command `runif(n = 1)`, you will get a random number between 0 and 1.

-   But if you run `set.seed(1)` before running `runif(n=1)` you will get the same result each time

-   This number is 0.2655087

## How to embed Reproducibility into your code

-   The most effective way to embed reproducibility into your code is to run `set.seed()` at the beginning of your script, before running that pipeline from end-to-end.

-   This means if someone was to run your code, with the exact same data, all cases of randomness will be defined in the same way.

-   We will be able to witness this in action shortly.

# Section 2: Introduction to Machine Learning

## Note:

As always, there are multiple packages which can be used to apply Machine Learning models.

Today we will focus on using `tidymodels`, which use a tidyverse approach.

However, other packages such as `e1071` or `rpart` are available for specific uses

## When to apply machine learning

Machine Learning problems can be seen as being part of three main groups:

-   To predict
-   To classify
-   To cluster, group or investigate

These each link to a different type of Machine Learning Method.

## Types of Machine Learning

Supervised Learning: Those which use *tagged* or *labelled* data to learn

-   Regression:
    -   Problem aim: Prediction
    -   Outcome Variable: Continuous
-   Classification:
    -   Problem aim: Classify
    -   Outcome variable: Categorical

Unsupervised Learning: Those which use *untagged* or *unlabeled* data to learn

-   Clustering:
    -   Problem aim: Grouping or Clustering
    -   Outcome variable: No specific variable type

## Matching variables to the methodology

When considering machine learning, the methodology will typically depend on the data available

-   Classification methods can only be used on Categorical outcome variables
-   Prediction methods can only be used on Continuous outcome variables

Of course, there is nuance in this, especially when you can convert continuous to categorical variables.

## Why Tidymodels

Tidymodels looks to apply the tidyverse style and readability to create a unified approach to building models. Taking advantage of the multitude of different packages each requiring different syntaxes and approaches.

This allows for easy scaling of models to include underwritters like Spark

## Building Models

To build effective models, you should take your data, and split it into:

-   **Training data** - This is where we develop our model \[Typically 80% of the data\]
-   **Testing data** - This is where we test the models accuracy \[Typically 20% of the data\]
-   **Validation data** (optional) - This is how we validate our findings

As this helps to reduce the risk of over fitting and improves models accuracy.

## How to split data

When splitting data, we typically want a 80:20 Training:Test split.

This can be achieved with tidymodels through the `initial_split()` function

-   `data_split <- initial_split(data, prop = 0.8)`
-   `data_training <- training(data_split)`
-   `data_test <- testing(data_split)`

To split for validation you can use:

-   `data_split <- initial_split(data, prop = c(0.6, 0.2))`
-   `data_valid <- validation(data_split)`

## Linking Reproducibility to Machine Learning

-   Being able to reproduce the splits you create in your data is vital.

-   You can control the randomness found, using `set.seed()` function

# Exercise 1: Applying Reproducibility, through Machine Learning Preparation

# Break

# Section 3: Developing Machine Learning Models

## Building a tidy model

Importantly, when using tidy models, we are required to *set an engine*, defining how the model should be developed for example:

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

# Classification Decision tree 
c_dt_model <- 
  decision_tree() |> 
  set_engine("rpart") |> 
  set_mode("classification")
```

In these cases you can also add the `translate()` function to convert the produced output into the same format as the traditional engine.

## Some Common & Accessible Machine Learning Methods

-   Supervised Learning
    -   Regression:
        -   Linear Regressions / GLM - `linear_reg()` with engines: `"lm"`
        -   Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"regression"`
        -   Support Vector Regression - `svm_linear()` with engines: `"kernlab"` and mode `"regression"`

## Some Common & Accessible Machine Learning Methods

-   Supervised Learning
    -   Classification:
        -   Logistic Regressions - `logistic_reg()` with engines: `"glm"`
        -   Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"classification"`
        -   Support Vector Machines - `svm_linear()` with engines: `"kernlab"` and mode `"classification"`

## Fitting Models

Once your model has been created, you can utilise the `fit()` function

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

```

You can then view the outcomes using:

```{r}
#| echo: TRUE
#| eval: false

model_res <- 
  lm_model_fit |>
  extract_fit_engine() |>
  summary()

```

## Making Predictions

You can then use this model, to make predictions using:

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

# Model Prediction 
lm_model_prediction <- 
  predict(lm_model_fit, 
          new_data = data_new)

# Bind the prediction with the "truth"
model_res <- bind_cols(lm_model_prediction, 
                       data_new |> select(x))

```

## Evaluating Models

This will create predictions based on the data provided, which can then be used to evaluate model effectiveness

```{r}
#| echo: TRUE
#| eval: false

# Set Model Effectiveness Measures (for linear regression)

model_metrics <- metric_set(rmse, rsq, mae)

# Apply the evaluation methods 

model_metrics(model_res, truth = x, estimate = .pred)

```

Please note, the model metrics, will vary depending on the method being used!

# Exercise 2: Applying Machine Learning Models

## Next Session:

-   Mini Project Preparation
-   23rd June, 12:30-14:00

## Resources:

-   [Tidy Modeling with R](https://www.tmwr.org/)
-   [Tidy Modeling Cheatsheet](https://www.tidymodels.org/find/parsnip/)
-   [Learning Statistics with R - Danielle Navarro](https://learningstatisticswithr.com/lsr-0.6.pdf)
-   [Introduction to Statistical Learning - James, Witten, Hastie & Tibshirani](https://www.statlearning.com/)
